### 2018-01-04 submission/manuscript.Rmd
methods re-write

***Obtaining Data Sets:***
Studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016]. Additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature. Studies that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing analysis were included. Only data sets that had sequences available for analysis were included. Some studies did not have publicly available sequences or did not have metadata in which the authors were able to share and were excluded. After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was not included because the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016]. Additionally, after sequence processing all the case samples for the Kostic study only had 100 or less sequences remaining and was not used. This left a total of 14 studies for which analysis could be completed.

@@

***Sequence Processing:***
For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained by searching the respective accession number of the study at the following website: http://www.ncbi.nlm.nih.gov/Traces/study/. Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap for one study [@ahn_human_2013] and for four studies was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]. Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009]. Where possible, quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing. If it was not possible to use these defaults, the stated quality cut-offs were used instead. Chimeras were identified and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011]. Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and variable region effects. Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many with and without disease were above and below the overall median value within the specific study. Relative risk significance testing utilized the chi-squared test. $\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017]. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002]. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages [@pwr_citation_2017; @statmod_citation_2016]. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages [@ggplot2_citation_2009; @gridextra_citation_2017].

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma. We analyzed the data using linear mixed-effect models and relative risk. $\beta$-diversity was then assessed for each individual study. Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk. We then built Random Forest models based on all genera or the select CRC-associated genera. The models were trained on one study then tested on the remaining studies for every study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched). Where applicable for each study, predictions for adenoma and carcinoma were then tested for feces, matched tissue, and unmatched tissue. This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold CV was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs. For OTU based models, the CRC-associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. Finally, the power of each study was assessed for an effect size ranging from 1% to 30%. An estimated sample size for these effect sizes was also generated based on 80% power. For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study. Similarly, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study.

***Obtaining CRC-Associated Genera:***
For the CRC-associated genera analysis of the RR, the total average counts were collected for each respective OTU that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas* for 100 different subsamplings. The OTU based Random Forest Models that used CRC-associated genera used a similar approach except that the OTUs were not aggregated together by genus by kept as separate OTUs. So, OTU Random Forest models using the full community included all OTUs while those using CRC-associated genera included only those OTUs that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.

***Matched versus Unmatched Tissue Samples:***
In general, tissue samples that had control and lesion samples that did not belong to the same individual were classified as unmatched while samples that belonged to the same individual were classified as matched. Studies with matched data included Burns, Dejea, Geng, and Lu. Studies that had unmatched data were Burns, Flemer, Chen, and Sanapareddy. For some studies samples became unmatched due to one of the corresponding matched samples not making it through sequence processing. For the linear mixed-effect models samples from the same individual were taken into account. For all other analysis matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.

***Assessing Important Random Forest Model Variables:***
The genus based models collected the top 10 most important variables, measured by Mean Decrease in Accuracy (MDA), from each training set and assessed how many times that genera showed up in the top 10 across each study. The OTU based models recorded the medians for each OTU across 100 different 80/20 splits of the data for each study. The lowest classification for each OTU was obtained using the RDP database and the number of times the specific classification occurred in the top 10 across studies was recorded. For the adenoma tissue genus and OTU models there was only one matched and unmatched study and these results were grouped together for the counting of the top 10.

***Reproducible Methods:***
The code and analysis can be found here https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017. Unless otherwise mentioned, the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.

>>>>

***Obtaining Data Sets:***
The studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016] and additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature. Studies that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing analysis were included and only data sets that had sequences available for analysis were included. Some studies were excluded because they did not have publicly available sequences or did not have metadata in which the authors were able to share and were excluded. After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014]. The Zackular [@zackular_human_2014] study was not included because the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016]. After sequence processing all the case samples for the Kostic study had 100 or less sequences remaining and was not used which left a total of 14 studies that analysis could be completed on.

@@

***Sequence Processing:***
For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained by searching the respective accession number of the study at the following website: http://www.ncbi.nlm.nih.gov/Traces/study/. Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap (n = 1, [@ahn_human_2013]) and directly from the authors (n = 4, [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]). Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009] and quality filtering utilized the default methods, used in mothur, for either 454 or Illumina based sequencing. If it was not possible to use the defaults, the stated quality cut-offs, from the study itself, were used instead. Chimeras were identified and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011]. Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hypervariable region used. Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study. Relative risk significance testing utilized the chi-squared test. $\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017]. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002]. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages [@pwr_citation_2017; @statmod_citation_2016]. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages [@ggplot2_citation_2009; @gridextra_citation_2017].

***Study Analysis Overview:***
$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma using both linear mixed-effect models and RR. $\beta$-diversity was then assessed for each individual study for differences between control-adenoma and control-carcinoma. Next, four specific mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in RR for adenoma and carcinoma. We then built Random Forest models based on all or mouth-associated genera and these models were trained on one study then tested on the remaining studies, for every study. A similar approach was then applied at the OTU level with the exception that a 10-fold CV over 100 different models, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs. For these OTU based models, the mouth-associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*. Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power. For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study. Similarly, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study. The data was split between feces and tissue samples. Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched). Where applicable for each study, predictions for adenoma and carcinoma were then tested for feces, matched tissue, and unmatched tissue.

***Obtaining Mouth-Associated Genera:***
For the mouth-associated genera analysis of the RR, the total average counts, for 100 different subsamplings, were collected for each respective OTU that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*. The OTU based Random Forest Models using mouth-associated genera utilized a similar approach except that the OTUs were not aggregated together by genus but kept as separate OTUs. OTU Random Forest models using the full community included all OTUs while those using mouth-associated genera included only those OTUs that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.

***Matched versus Unmatched Tissue Samples:***
In general, tissue samples that had control and lesion samples, that did not belong to the same individual, were classified as unmatched while samples, that belonged to the same individual, were classified as matched. Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy. For some studies samples became unmatched due to one of the corresponding matched samples not making it through sequence processing. All samples, from both tissue sample types, were analyzed together for the linear mixed-effect models with samples from the same individual corrected for. For all other analysis, not mentioned herein, matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.

***Assessing Important Random Forest Model Variables:***
Using Mean Decrease in Accuracy (MDA) the top 10 most important variables were obtained in two different ways depending on whether the model used genera or OTU data. For the genus based models, the number of times that a genus showed up in the top 10 of the training set across each study was counted while, for the OTU based models, the medians for each OTU across 100 different 80/20 splits of the data was generated and the top 10 OTUs then counted for each study. Common taxa, for the OTU based models, were identified by using the lowest classification within the RDP database for each of the specific OTUs obtained from the previous counts and the number of times this classification occurred in the top 10, in each study, was recorded. The two studies that had adenoma tissue were equally divided between matched and unmatched groups and were grouped together for the counting of the top 10 genera and OTUs.

***Reproducible Methods:***
The code and analysis can be found here at  https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017. Unless otherwise mentioned, the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.

### 2018-01-04 submission/manuscript.tex
methods re-write

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, \emph{et
al.} and Vogtmann, \emph{et al.} {[}29,30{]}. Additional studies not
mentioned in the reviews were obtained based on the authors' knowledge
of the literature. Studies that used tissue or feces as their sample
source for 454 or Illumina 16S rRNA gene sequencing analysis were
included. Only data sets that had sequences available for analysis were
included. Some studies did not have publicly available sequences or did
not have metadata in which the authors were able to share and were
excluded. After these filtering steps, the following studies remained:
Ahn, \emph{et al.} {[}26{]}, Baxter, \emph{et al.} {[}16{]}, Brim,
\emph{et al.} {[}22{]}, Burns, \emph{et al.} {[}27{]}, Chen, \emph{et
al.} {[}19{]}, Dejea, \emph{et al.} {[}24{]}, Flemer, \emph{et al.}
{[}12{]}, Geng, \emph{et al.} {[}28{]}, Hale, \emph{et al.} {[}18{]},
Kostic, \emph{et al.} {[}15{]}, Lu, \emph{et al.} {[}21{]}, Sanapareddy,
\emph{et al.} {[}25{]}, Wang, \emph{et al.} {[}20{]}, Weir, \emph{et
al.} {[}23{]}, and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}31{]}
study was not included because the 90 individuals analyzed within the
study are contained within the larger Baxter study {[}16{]}.
Additionally, after sequence processing all the case samples for the
Kostic study only had 100 or less sequences remaining and was not used.
This left a total of 14 studies for which analysis could be completed.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
for one study {[}26{]} and for four studies was obtained directly from
the authors {[}12,18,23,25{]}. Each study was processed using the mothur
(v1.39.3) software program {[}32{]}. Where possible, quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults, the
stated quality cut-offs were used instead. Chimeras were identified and
removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}34{]}.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.2) software package {[}35{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}36{]} and then Z-score
normalized using the car (v2.1.5) package {[}37{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}38{]} to correct for study,
repeat sampling of individuals (tissue only), and variable region
effects. Relative risk was analyzed using both the epiR (v0.9.87) and
metafor (v2.0.0) packages {[}39,40{]} by assessing how many with and
without disease were above and below the overall median value within the
specific study. Relative risk significance testing utilized the
chi-squared test. \(\beta\)-diversity differences utilized a Bray-Curtis
distance matrix and PERMANOVA executed with the vegan (v2.4.4) package
{[}41{]}. Random Forest models were built using both the caret (v6.0.77)
and randomForest (v4.6.12) packages {[}42,43{]}. Power analysis and
estimations were made using the pwr (v1.2.1) and statmod (v1.4.30)
packages {[}44,45{]}. All figures were created using both ggplot2
(v2.2.1) and gridExtra (v2.3) packages {[}46,47{]}.

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma. We
analyzed the data using linear mixed-effect models and relative risk.
\(\beta\)-diversity was then assessed for each individual study. Next,
four specific CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas})
were assessed for differences in relative risk. We then built Random
Forest models based on all genera or the select CRC-associated genera.
The models were trained on one study then tested on the remaining
studies for every study. The data was split between feces and tissue
samples. Within the tissue groups the data was further divided between
samples from the same individual (matched) and those from different
individuals (unmatched). Where applicable for each study, predictions
for adenoma and carcinoma were then tested for feces, matched tissue,
and unmatched tissue. This same approach was then applied at the OTU
level with the exception that instead of testing on the other studies, a
10-fold CV was utilized and 100 different models were created based on
random 80/20 splitting of the data to generate a range of expected AUCs.
For OTU based models, the CRC-associated genera included all OTUs that
had a taxonomic classification to \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas}.
Finally, the power of each study was assessed for an effect size ranging
from 1\% to 30\%. An estimated sample size for these effect sizes was
also generated based on 80\% power. For comparisons in which normal
versus adenoma were made the carcinoma samples were excluded from each
respective study. Similarly, for comparisons in which normal versus
carcinoma were made the adenoma samples were excluded from each
respective study.

\textbf{\emph{Obtaining CRC-Associated Genera:}} For the CRC-associated
genera analysis of the RR, the total average counts were collected for
each respective OTU that had a genus level taxonomic classification to
\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, and
\emph{Porphyromonas} for 100 different subsamplings. The OTU based
Random Forest Models that used CRC-associated genera used a similar
approach except that the OTUs were not aggregated together by genus by
kept as separate OTUs. So, OTU Random Forest models using the full
community included all OTUs while those using CRC-associated genera
included only those OTUs that had a genus level taxonomic classification
to \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus},
and \emph{Porphyromonas}.

\textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
tissue samples that had control and lesion samples that did not belong
to the same individual were classified as unmatched while samples that
belonged to the same individual were classified as matched. Studies with
matched data included Burns, Dejea, Geng, and Lu. Studies that had
unmatched data were Burns, Flemer, Chen, and Sanapareddy. For some
studies samples became unmatched due to one of the corresponding matched
samples not making it through sequence processing. For the linear
mixed-effect models samples from the same individual were taken into
account. For all other analysis matched and unmatched samples were
analyzed separately using the statistical approaches mentioned in the
Statistical Analysis section.

\textbf{\emph{Assessing Important Random Forest Model Variables:}} The
genus based models collected the top 10 most important variables,
measured by Mean Decrease in Accuracy (MDA), from each training set and
assessed how many times that genera showed up in the top 10 across each
study. The OTU based models recorded the medians for each OTU across 100
different 80/20 splits of the data for each study. The lowest
classification for each OTU was obtained using the RDP database and the
number of times the specific classification occurred in the top 10
across studies was recorded. For the adenoma tissue genus and OTU models
there was only one matched and unmatched study and these results were
grouped together for the counting of the top 10.

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
here
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Unless otherwise mentioned, the accession number for the raw sequences
for the studies used in this analysis can be found directly in the
respective batch file in the GitHub repository or in the original
manuscript.

>>>>

\textbf{\emph{Obtaining Data Sets:}} The studies used for this
meta-analysis were identified through the review articles written by
Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}29,30{]} and
additional studies not mentioned in the reviews were obtained based on
the authors' knowledge of the literature. Studies that used tissue or
feces as their sample source for 454 or Illumina 16S rRNA gene
sequencing analysis were included and only data sets that had sequences
available for analysis were included. Some studies were excluded because
they did not have publicly available sequences or did not have metadata
in which the authors were able to share and were excluded. After these
filtering steps, the following studies remained: Ahn, \emph{et al.}
{[}26{]}, Baxter, \emph{et al.} {[}16{]}, Brim, \emph{et al.} {[}22{]},
Burns, \emph{et al.} {[}27{]}, Chen, \emph{et al.} {[}19{]}, Dejea,
\emph{et al.} {[}24{]}, Flemer, \emph{et al.} {[}12{]}, Geng, \emph{et
al.} {[}28{]}, Hale, \emph{et al.} {[}18{]}, Kostic, \emph{et al.}
{[}15{]}, Lu, \emph{et al.} {[}21{]}, Sanapareddy, \emph{et al.}
{[}25{]}, Wang, \emph{et al.} {[}20{]}, Weir, \emph{et al.} {[}23{]},
and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}31{]} study was not
included because the 90 individuals analyzed within the study are
contained within the larger Baxter study {[}16{]}. After sequence
processing all the case samples for the Kostic study had 100 or less
sequences remaining and was not used which left a total of 14 studies
that analysis could be completed on.

@@

\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
sequences were downloaded from the Sequence Read Archive (SRA)
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained by searching the respective accession number
of the study at the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
not have sequences and metadata on the SRA, data was obtained from DBGap
(n = 1, {[}26{]}) and directly from the authors (n = 4,
{[}12,18,23,25{]}). Each study was processed using the mothur (v1.39.3)
software program {[}32{]} and quality filtering utilized the default
methods, used in mothur, for either 454 or Illumina based sequencing. If
it was not possible to use the defaults, the stated quality cut-offs,
from the study itself, were used instead. Chimeras were identified and
removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
97\% similarity was completed using the OptiClust algorithm {[}34{]}.

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}35{]}.
For the \(\alpha\)-diversity analysis, values were power transformed
using the rcompanion (v1.10.1) package {[}36{]} and then Z-score
normalized using the car (v2.1.5) package {[}37{]}. Testing for
\(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.14) package {[}38{]} to correct for study,
repeat sampling of individuals (tissue only), and 16S hypervariable
region used. Relative risk was analyzed using both the epiR (v0.9.87)
and metafor (v2.0.0) packages {[}39,40{]} by assessing how many
individuals with and without disease were above and below the overall
median value within each specific study. Relative risk significance
testing utilized the chi-squared test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.4) package {[}41{]}. Random Forest models were built using
both the caret (v6.0.77) and randomForest (v4.6.12) packages
{[}42,43{]}. Power analysis and estimations were made using the pwr
(v1.2.1) and statmod (v1.4.30) packages {[}44,45{]}. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
{[}46,47{]}.

\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
assessed for differences between controls, adenoma, and carcinoma using
both linear mixed-effect models and RR. \(\beta\)-diversity was then
assessed for each individual study for differences between
control-adenoma and control-carcinoma. Next, four specific
mouth-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
differences in RR for adenoma and carcinoma. We then built Random Forest
models based on all or mouth-associated genera and these models were
trained on one study then tested on the remaining studies, for every
study. A similar approach was then applied at the OTU level with the
exception that a 10-fold CV over 100 different models, based on random
80/20 splitting of the data, was used to generate a range of expected
AUCs. For these OTU based models, the mouth-associated genera included
all OTUs that had a taxonomic classification to \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas}.
Finally, the power of each study was assessed for an effect size ranging
from 1\% to 30\% and an estimated sample size, for these effect sizes,
was generated based on 80\% power. For comparisons in which normal
versus adenoma were made the carcinoma samples were excluded from each
respective study. Similarly, for comparisons in which normal versus
carcinoma were made the adenoma samples were excluded from each
respective study. The data was split between feces and tissue samples.
Within the tissue groups the data was further divided between samples
from the same individual (matched) and those from different individuals
(unmatched). Where applicable for each study, predictions for adenoma
and carcinoma were then tested for feces, matched tissue, and unmatched
tissue.

\textbf{\emph{Obtaining Mouth-Associated Genera:}} For the
mouth-associated genera analysis of the RR, the total average counts,
for 100 different subsamplings, were collected for each respective OTU
that had a genus level taxonomic classification to \emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}.
The OTU based Random Forest Models using mouth-associated genera
utilized a similar approach except that the OTUs were not aggregated
together by genus but kept as separate OTUs. OTU Random Forest models
using the full community included all OTUs while those using
mouth-associated genera included only those OTUs that had a genus level
taxonomic classification to \emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, and \emph{Porphyromonas}.

\textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
tissue samples that had control and lesion samples, that did not belong
to the same individual, were classified as unmatched while samples, that
belonged to the same individual, were classified as matched. Studies
with matched data included Burns, Dejea, Geng, and Lu while those with
unmatched data were from Burns, Flemer, Chen, and Sanapareddy. For some
studies samples became unmatched due to one of the corresponding matched
samples not making it through sequence processing. All samples, from
both tissue sample types, were analyzed together for the linear
mixed-effect models with samples from the same individual corrected for.
For all other analysis, not mentioned herein, matched and unmatched
samples were analyzed separately using the statistical approaches
mentioned in the Statistical Analysis section.

\textbf{\emph{Assessing Important Random Forest Model Variables:}} Using
Mean Decrease in Accuracy (MDA) the top 10 most important variables were
obtained in two different ways depending on whether the model used
genera or OTU data. For the genus based models, the number of times that
a genus showed up in the top 10 of the training set across each study
was counted while, for the OTU based models, the medians for each OTU
across 100 different 80/20 splits of the data was generated and the top
10 OTUs then counted for each study. Common taxa, for the OTU based
models, were identified by using the lowest classification within the
RDP database for each of the specific OTUs obtained from the previous
counts and the number of times this classification occurred in the top
10, in each study, was recorded. The two studies that had adenoma tissue
were equally divided between matched and unmatched groups and were
grouped together for the counting of the top 10 genera and OTUs.

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
here at
\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
Unless otherwise mentioned, the accession number for the raw sequences
for the studies used in this analysis can be found directly in the
respective batch file in the GitHub repository or in the original
manuscript.
